Introduction to *cellcount* - Draft vignette

### 2023-04-26 cellcount_intro.Rmd
Updates made - initial changes from author review and overall changes to third person narration.

---
title: "Introduction to *cellcount* - Draft vignette"
author: "*T.E. Harman, M. Vandersea, K. Pokrzywinski*"
date: "04/19/2023"
abstract: >
  The *cellcount* package provides wrapper functions for  [*EBImage*](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html) to simplfy quantifying bacteria and cyanobacteria via flourescence microscopy. Cyanobacteria enumeration is an important tool to incorporate data results into mathematical models to assess bloom dynamics. However, standard manual cyanobacteria enumeration is considered time-consuming and can impact overall random error due to different inaccuracies from researcher to researcher. Here, we outline a new open-source tool, *cellcount*, a package designed for the computing language R to assist with cyanobacteria enumeration. Fluorescent microscopy images are analyzed by this package to generate cyanobacteria counts to assess overall cell density from samples.
output: html_document
---

@@

*cellcount* is an R package distributed freely by NOAA NCCOS HAB-F on [GitHub](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories) and uses components in the RStudio IDE program. Make sure to install both the latest version of R and RStudio.

To get started with *cellcount*, users can make a local clone of the repository using programs like [GitHub Desktop](https://desktop.github.com/) or similar programs. Users may also download the package from the provided release on GitHub that contains a zipped folder with the package material. Once cloned, open the R project file and install the package locally by selecting 'build' and 'install package', or use the shortcut Crtl+Shift+B (Cmd+Shift+B for MacOS users). You should now have the *cellcount* package installed. Use the following code to bring *cellcount* into the R environment:

@@

In addition, the following packages are required to use functions within *cellcount*:

- *EBImage*
- *tiff*
- *pixmap*
- *raster*
- *beepr*

*EBImage* uses the [Bioconductor](http://bioconductor.org) development software system while the remaining packages use the CRAN server built within R. Use the following code to install and load the packages into the R environment before starting:

@@

Now that we have the basics installed, we can begin using the *cellcount* package. Start by generating a data frame for count data to be added into using the following code:

@@

You'll need to generate a couple of directories; one for saving the unique
color label images and another for CSV files. In this example, we are using 
images *Microcystis* sp. cells. Please note that your 
directory addresses will be different. We recommend using a dedicated directory 
for *cellcount* results:

@@

Next, you'll need to read in the images you are interested in processing using
the following code. Once again, your directories will be different and must be
updated to the correct file address. However, in this example we will be using
the provided test images from this package:

```{r,message=FALSE}
images <- list.files("C:/file path of images to analyze", pattern = "tif", full.name = T)
images_names <- list.files("C:/file path of images to analyze", pattern = "tif", full.name = F)
```
```{r,include=FALSE}
images <- list.files("./inst/extdata/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = T)
images_names <- list.files("./inst/extdata/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = F)
```

In addition, you'll need to assign the file names to a vector along with adding
in characterization to each file for easy organization:

@@

Now we will "read" in the images that we are interested in analyzing. The
following code outlines this step using the "readTIFF" function from the 
*tiff* package. In addition, we'll need to re-orient the image as it's uploaded
into the environment. The "readTIFF" function uploads images in a portrait 
format - the "aperm" function reorients the uploaded image to a landscape format:

@@

Now that our images are within the R environment, we can begin to convert these
images into useful data that R can use to process images. First, we begin by
converting 24-bit images into 8-bit images using the "greyscale" function (see 
photo below). This function also has a built-in numerical contrast adjustment 
that end-users can adjust:

@@

Next, we will use the "mapped" function to turn the 8-bit image we just
converted into a data matrix. In addition, we use an additional function to 
determine the median intensity of each image. From this, users can add a numerical
buffer on top of the determined median intensity, giving a more accurate approach
to separating the background intensity from the intensity of cells within each
image:

@@

Before we analyze our images in bulk, set aside some code space to visually 
analyze your images results. This could include adjusting the threshold
from the "mapped" function, the contrast from the "greyscale" function 
(if used), or could involve making adjustments to the "single_cell_convert" or
"filamentous_convert" functions, such as the width and height of the rectangular 
window, the pixel area threshold (i.e., removes cells or other objects under a 
certain pixel area),or the tolerance and 'ext' variables. Observe your initial 
results as indicated below and make adjustments where necessary:

@@

Once you have made necessary adjustments, use the following for-loop to
analyze your images in bulk, one at a time. In this example using *Microcystis* sp.
images, the for-loop uses the "single_cell_convert" function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the "count_images" function. Finally, "count_cells" is used 
to count the number of unique color labels (i.e., cells). We also have 
integrated code to add these count results to the data frame made earlier, and 
also to save each unique color label image to the "image_savdir" directory:

@@

Data from the image analysis has been generated in our data frame "Cell.Count"
that we previously made. We can use the following code to view our results:

@@

Some images that were initially taken may be too poor of quality or contain
high background fluorescence, among other errors. The code below analyzes 
outliers from the produced data, within a range of two standard deviations 
above and below the median. The "outlier_detect" function tags which images are
out of this range and dictates a warning within our data frame, which is shown 
below. This alerts the end-user to perhaps re-image a particular sample, or
to investigate the error further.

@@

In order to calculate cell density of these images, we can use the
"cell_density" function that's built into the *cellcount* package. We first
need to generate the total amount of cells counted from these images and
add it to our existing data frame using the following code:

@@

*Important note* - *the "null" data produced in the code above (and others that*
*follow) fills in the blanks from the previous data produced by the* 
*"outlier" function.*

Calculating cell density, taken from Wetzel and Likens 1991, requires multiple
values. End-users will need the following information to precisely 
calculate this value:

@@

With the variables calculated, the following code uses these variables to 
calculate cell density and adds the data into the "Cell.Count" data frame:

@@

We can also calculate the average number of cells per image, as well as cells
per mL using the following code. Note that 'cell.av' must be divided by the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

@@

We can now view our data frame to double check for any calculation errors:

@@

Finally, we write the following code to export our finalized .csv file to our
pre-determined save directory:

@@

# Output data from test images in the *cellcount* package:

@@

Several informal discussions with researchers have expressed interest in a easy-to-use analysis interface for inexperienced R users
to utilize the ***cellcount*** tool. The following overview uses ***Shiny*** apps to enter input data and run image analyses.

### Inputting data
The *input_data_GUI* function generates a Shiny UI where users enter information regarding directories, numerical and categorical data, 
and selecting analysis type. Tabs to the left of the UI separate particular categories for easy organization. Once all data has been
provided, users complete the input of data by selecting the "Completion" tab and clicking the "Submit data entries" button to export 
data to the local R environment.

@@

### Running the image analysis

*analyze_images_GUI* is a simple UI that contains three interactive buttons as well as a data table below. "Run Image Analysis" simply runs the image analysis previously outlined in this document and uses the inputted data from the *input_data_GUI* function. Once the completion sound has been played (using the *beepr* package), users can click the "Refresh" button to update the table below with the analyzed data and review. Once finished with analysis, the "Close window" button to exit the UI.

@@

The current version of this package may be limiting for some users regarding processing power. Analysis with this package uses local memory, which may be limiting if expected to analyze large amounts of images in batches. We recommend at least 16GB of localized RAM, the use of a cloud computing supercomputer (i.e. Microsoft Azure), or organize images in easier-to-analyze batches to reduce local memory strain.

Users must be aware that this package does not contain abilities to identify distinct cells. The development of this package utilized carefully cultured monoculture of various cyanobacteria species. Users may find difficulty in using this analysis tool if cultures contain heavy amounts of particulates, other cellular debris, and contamination of other species. It is advised to use this tool for single-species only, however future developments of this tool are planned include identifying cell features and building image libraries of specific cellular species for use in environmental samples.

Outside of package limitations, users must be aware of limitations that may arise from poor microscopy imaging; reduce lighting as much as possible, avoid heavy background fluorescence, reduce debris/particulates, and so forth. Good images produce good *cellcount* results - keep this in mind as you utilize this tool.

>>>>

---
title: "Introduction to *cellcount* - Draft vignette"
author: "*T.E. Harman, M. Vandersea, K. Pokrzywinski*"
date: "`r Sys.Date()`"
abstract: >
  The ***cellcount*** package provides a simple approach to quantifying bacteria and cyanobacteria *via* flourescence microscopy. Microbial enumeration is an important tool for validating new methods and critical to assess potential ecological changes. However, standard manual microbial enumeration is time-consuming and notoriously difficult due to large colonies, filaments, and dense biomats. Conventional counting methods require specialized expertise and are error prone due to variations between researchers. Additionally, other methods of counting cells or particles (i.e. coulter counter, flow cytometer, etc.) are not suitable for filamentous or colonial morphotypes that can clog flow cells and apertures. Here, we outline a new open-source package, ***cellcount***, designed in the R computing language, to quickly and accurately count various microbial morphotypes. In this open access package, fluorescent microscope images are analyzed to generate particle counts and quantify overall cell density from samples at comparable concentrations as conventional counting methods.
output: html_document
---

@@

***cellcount*** is an open-source R package distributed by NOAA NCCOS HAB-F on [GitHub](https://github.com/orgs/HABF-SDI-NCCOS-NOAA/repositories) and uses components in the RStudio IDE program. It is advised to use both the latest version of R and RStudio before beginning.

To get started with ***cellcount***, users can make a local clone of the repository using programs like [GitHub Desktop](https://desktop.github.com/) or similar programs. Users may also download the package from the provided release on GitHub that contains a zipped folder with the package material. Once cloned, open the R project file and install the package locally by selecting 'build' and 'install package', or use the shortcut Crtl+Shift+B (Cmd+Shift+B for MacOS users). Once the ***cellcount*** package is installed, the following code can be used to bring ***cellcount*** into the R environment:

@@

In addition, the following packages are required to use functions within ***cellcount***:

- [***EBImage***](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html)
- [***tiff***](https://www.rforge.net/tiff/)
- [***pixmap***](https://cran.r-project.org/package=pixmap)
- [***raster***](https://cran.r-project.org/web/packages/raster/index.html)
- [***beepr***](https://cran.r-project.org/web/packages/beepr/index.html)

***EBImage*** uses the [Bioconductor](http://bioconductor.org) development software system while the remaining packages use the CRAN server built within R. Use the following code to install and load the packages into the R environment before starting:

@@

After the basic supporting packages have been installed, an empty data frame can be generated to record resulting count data from the ***cellcount*** package using the following code:

@@

In order to appropriately save files from the ***cellcount*** analysis, some directories will be required; one for saving the unique color label images and another for CSV files. Please note that directory addresses will be different based on operating system and location. It is recommended to use a dedicated directory for ***cellcount*** results:

@@

Next, images that are of interest will need to read into the R environment using
the following code. Once again, directories will be different and must be
updated to the correct file address. This example will use 'tif' images of *Microcystis* sp. that are provided within the ***cellcount*** package:

```{r}
images <- list.files("./inst/extdata/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = T)
images_names <- list.files("./inst/extdata/test_images/Microcystis_F192_40x/", pattern = "tif", full.name = F)
```

In addition, users can assign additional identifications to each file for easy organization, as shown here:

@@

Images from the assigned directory can now be "read" into the environment for analysis. The
following code outlines this step using the *readTIFF* function from the 
***tiff*** package. In addition, images will need to be re-oriented as the *readTIFF* function uploads images in a portrait 
format. The *aperm* function shown below re-orients the uploaded image to a landscape format:

@@

Once images are within the R environment, the process to convert these
images into useful data can begin. First, begin by
converting 24-bit images into 8-bit images using the *greyscale* function (see 
photo below). This function also has a built-in numerical contrast adjustment 
that end-users can change:

@@

The *mapped* function turns the converted 8-bit image into a data matrix. An additional function to 
determine the median intensity of each image is also used. From this, users can add a numerical
buffer on top of the determined median intensity, giving a more accurate approach
to separating the background intensity from the intensity of cells within each
image. For example, if an image has a median intensity of 0.05, with a numerical buffer of 0.2,
the threshold is equal to 0.25. Any pixel within the image below 0.25 will be omitted from the matrix and assigned a value of zero.

@@

Before images are analyzed in bulk, it is encouraged to visually 
analyze image results and determine if changes need to be made. This could include adjusting the threshold
from the "mapped" function, the contrast from the "greyscale" function 
(if used), or could involve making adjustments to the "single_cell_convert" or
"filamentous_convert" functions, such as the width and height of the rectangular 
window, the pixel area threshold (i.e., removes cells or other objects under a 
certain pixel area), or the tolerance and 'ext' variables. Observe initial 
results as indicated below and make adjustments where necessary:

@@

Once necessary adjustments have been made, the following for-loop is used to
analyze images of interested in bulk, one at a time. In this example, using the provided demo *Microcystis* sp.
images, the for-loop uses the "single_cell_convert" function to produce unique 
color labels on cells within an image and to produce a final image showcasing 
these labels using the "count_images" function. Finally, "count_cells" is used 
to count the number of unique color labels (i.e., cells). Code has also been integrated in the for-loop
to add these count results to the 'Cell.Count' data frame made earlier, and 
to save each unique color label image to the "image_savdir" directory:

@@

Data from the image analysis has been written to the data frame "Cell.Count", where
calling the data frame can view these results:

@@

In same cases, some images that were initially taken may be too poor of quality or contain
high background fluorescence, among other errors that may arise. The following code analyzes 
outliers from the produced data, within a range of two standard deviations 
above and below the median. The "outlier_detect" function tags which images are
out of this range and dictates a warning within our data frame, which is shown 
below. This alerts the end-user to perhaps re-image a particular sample, or
to investigate the error further:

@@

In order to calculate cell density of these images, the
"cell_density" function can be implemented, using equations from Wetzel and Likens, 1991. Generating the total amount of 
cells counted from these images is needed, where it can be added to the 'Cell.Count' data frame using the following code:

@@

*Important note* - *the "null" data produced in the code above (and others that*
*follow) fills in the blanks from the previous data produced by the* 
***outlier_detect*** *function.*

Calculating cell density requires multiple
values. End-users will need the following information to precisely 
calculate this value:

@@

With the variables calculated, the following code uses these variables to 
calculate cell density and adds the data into the 'Cell.Count' data frame:

@@

The average number of cells per image can also be calculated, as well as cells
per mL using the following code. Note that 'cell.av' must be divided by the
number of images the end-user has analyzed, and the 'cell.mL' must be divided
by the total volume from the original sample:

@@

Calling upon the 'Cell.Count' data frame can show the new generated data and to double check for any calculation errors:

@@

Finally, use the following code to export the finalized data frame to a .csv file to the
pre-determined save directory:

@@

# Output data from test images in the ***cellcount*** package:

@@

Informal discussions with researchers have expressed interest in a easy-to-use analysis interface for inexperienced R users
to utilize the ***cellcount*** tool. The following overview uses ***shiny*** apps to enter input data and run image analyses. Several packages are required to utilize these UI tools:

- [***shiny***](https://cran.r-project.org/web/packages/shiny/index.html)
- [***shinyalert***](https://cran.r-project.org/web/packages/shinyalert/index.html)
- [***shinyFiles***](https://cran.r-project.org/web/packages/shinyFiles/index.html)
- [***shinyjs***](https://cran.r-project.org/web/packages/shinyjs/index.html)
- [***DT***](https://cran.r-project.org/web/packages/DT/index.html)
- [***shinyBS***](https://cran.r-project.org/web/packages/shinyBS/index.html)

Users can use the following code to bring packages into the R environment once the packages have been installed using CRAN:

```{r,message=FALSE,warning=FALSE}
library(shiny)
library(shinyalert)
library(shinyFiles)
library(shinyjs)
library(DT)
library(shinyBS)
```

### Inputting data
The *input_data_GUI* function generates a Shiny UI where users enter information regarding directories, numerical and categorical data, 
and selecting analysis type. Tabs to the left of the UI separate particular categories for easy organization and contain interactive buttons or drop-down menus to enter and select appropriate data. Some buttons change from red to green to provide an organized data entry approach. Once all data has been provided, users complete the input of data by selecting the "Completion" tab and clicking the "Submit data entries" button to export data to the local R environment.

@@

### Running the image analysis
*analyze_images_GUI* is a simple UI that contains three interactive buttons as well as an interactive data table. "Run Image Analysis" simply runs the image analysis previously outlined in this document and uses the inputted data from the *input_data_GUI* function. Once the completion sound has been played (using the ***beepr*** package), users can click the "Refresh" button to update the table below with the analyzed data and review. Once finished with analysis, the "Close window" button to exit the UI.

@@

The current version of this package may be limiting for some users regarding processing power. Analysis with this package uses local memory, which may be limiting if expected to analyze large amounts of images in batches. It is recommended to have at least 16GB of localized RAM, the ability to use cloud computing (i.e. Microsoft Azure), or organize images in easier-to-analyze batches to reduce local memory strain.

Users must be aware that this package does not contain abilities to identify distinct cells. The development of this package utilized monoculture of various cyanobacteria species. Users may find difficulty in using this analysis tool if cultures contain heavy amounts of particulates, other cellular debris, and contamination of other species. It is advised to use this tool for single-species only, however future developments of this tool are planned to include identifying cell features and deep-learning mechanics that use image libraries of specific cellular species.

Outside of package limitations, users must be aware of limitations that may arise from poor microscopy imaging; reduce lighting as much as possible, avoid heavy background fluorescence, reduce debris/particulates, and so forth. Good images produce good ***cellcount*** results - keep this in mind as you utilize this tool.

# Conclusions:
